

\begin{lemma}[Maehara]
	\label{lemma:maehara}
	Let $\Gamma$ and $\Delta$ be sets of first-order formulas without equality and function symbols such that $\Gamma \vdash \Delta$ is provable in cut-free sequent calculus.
	Then for any partition \parti{\Gamma_1}{\Delta_1}{\Gamma_2}{\Delta_2}
	there is an interpolant $I$ such that
	\begin{compactenum}
	\item $\Gamma_1 \proves \Delta_1, I$ is provable 
		\label{maehcond1}
	\item $\Gamma_2, I \proves \Delta_2$ is provable 
		\label{maehcond2}
	\item $\Lang(I) \subseteq \Lang(\Gamma_1, \Delta_1) \cap \Lang(\Gamma_2, \Delta_2)$
		\label{maehcond3}
	\end{compactenum}
\end{lemma}
\begin{proof}
	We prove this lemma by induction on the number of inferences in a cut-free proof of $\Gamma \proves \Delta$.
	By Lemma \ref{lemma:no_equality_in_proof}, we can assume that no equality symbol occurs in the proof, so equality rules need not be considered. 
	\begin{description}
		\item[\normalfont Base case.]
			Suppose no rules were applied.
			Then $C \vdash D$ is of one of the form
				$A \vdash A$. We give interpolants for any of the four possible partitions:
					\begin{enumerate}
						\item \parti{A}{A}{}{}: $I=\bot$
						\item \parti{}{}{A}{A}: $I=\top$
						\item \parti{}{A}{A}{}: $I=\lnot A$\nopagebreak
						\item \parti{A}{}{}{A}: $I=A$
					\end{enumerate}

		\item[\normalfont Structural rules.]
			Suppose the property holds for $n$ rule applications and the $(n+1)$th rule application is a structural one.

			\begin{itemize}
				\item The last rule application is an instance of \lkrule{c}{l}. Then it is of the form:
					\begin{prooftree}
						\Axiomm{\Gamma, A, A \fCenter \Delta}
						\RightLabelm{\lkrule{c}{l}}
						\UnaryInfm{\Gamma, A \fCenter \Delta}
					\end{prooftree}

					There are two possible partition schemes: of $\Gamma, A \proves \Delta$:
					\begin{enumerate}
						\item $\partisym = \parti{\Gamma_1, A}{\Delta_1}{\Gamma_2}{\Delta_2}$.
							By the induction hypothesis, we know that there is an interpolant $I$ for the partition \parti{\Gamma_1, A, A}{\Delta_1}{\Gamma_2}{\Delta_2} of the upper sequent.
							$I$ serves as interpolant for $\partisym$ as well.

						\item $\partisym = \parti{\Gamma_1}{\Delta_1}{\Gamma_2, A}{\Delta_2}$.
							By a similar argument, we get that there is an interpolant $I$ for 
							\parti{\Gamma_1}{\Delta_1}{\Gamma_2, A, A}{\Delta_2}, which again is also an interpolant for $\partisym$.

					\end{enumerate}

					The case of \lkrule{c}{r} is analogous.

				\item The last rule application is an instance of \lkrule{w}{r}. Then it is of the form:
					\begin{prooftree}
						\Axiomm{\Gamma \fCenter \Delta}
						\RightLabelm{\lkrule{w}{r}}
						\UnaryInfm{\Gamma \fCenter \Delta, A}
					\end{prooftree}

					By the induction hypothesis, there exists an interpolant $I$ for any partition \parti{\Gamma_1}{\Delta_1}{\Gamma_2}{\Delta_2} of $\Gamma \vdash \Delta$.
					Clearly $I$ remains an interpolant when adding $A$ to either $\Delta_1$ or $\Delta_2$.

					The case of \lkrule{w}{l} is analogous.

			\end{itemize}

		\item[\normalfont Propositional rules.]
			Suppose the property holds for $n$ rule applications and the $(n+\nolinebreak 1)$th rule application is a propositional one.

			\begin{itemize}
				\item The last rule application is an instance of \lkrule{\lnot}{l}. Then it is of the form:
					\begin{prooftree}
						\Axiomm{\Gamma \fCenter \Delta,  A}
						\RightLabelm{\lkrule{\lnot}{l}}
						\UnaryInfm{\lnot A, \Gamma \fCenter \Delta }
					\end{prooftree}

					There are two possible partition schemes of $\Gamma, \lnot A \vdash \Delta$:
					\begin{enumerate}
						\item $\partisym = \parti{\Gamma_1, \lnot A}{\Delta_1}{\Gamma_2}{\Delta_2}$.
							By the induction hypothesis, there exists an interpolant $I$ for the partition \parti{\Gamma_1}{\Delta_1, A}{\Gamma_2}{\Delta_2} of the upper sequent.
							Clearly $I$ is an interpolant for $\partisym$ as well.

						\item $\partisym = \parti{\Gamma_1}{\Delta_1}{\Gamma_2, \lnot A}{\Delta_2}$. A similar argument goes through. 
					\end{enumerate}

					The case of \lkrule{\lnot}{r} is analogous.

				\item The last rule application is an instance of \lkrule{\limpl}{l}. Then it is of the form:
					\begin{prooftree}
						\Axiomm{\Gamma \fCenter \Delta,  A}
						\Axiomm{\Sigma, B \fCenter \Pi}
						\RightLabelm{\lkrule{\limpl}{l}}
						\BinaryInfm{\Gamma, \Sigma, A \limpl B \fCenter \Delta, \Pi }
					\end{prooftree}

					There are two possible partition schemes of $\Gamma, A\limpl B \vdash \Delta$:
					\begin{enumerate}
						\item $\partisym = \parti{\Gamma_1, \Sigma_1, A\limpl B}{\Delta_1, \Pi_1}{\Gamma_2, \Sigma_2}{\Delta_2, \Pi_2}$.
							By the induction hypothesis, there is an interpolant $I_1$ for the partition $\parti{\Gamma_1}{\Delta_1, A}{\Gamma_2}{\Delta_2}$ of the left upper sequent.
							Hence for $I_1$, we have that $\Gamma_1 \fCenter \Delta_1, A, I_1$ and 
							$I_1, \Gamma_2 \fCenter \Delta_2$ are provable.

							Moreover, we also get by the induction hypothesis that there is an interpolant $I_2$ for the partition $\parti{\Sigma_1, B}{\Pi_1}{\Sigma_2}{\Pi_2}$ of the right upper sequent.
							Therefore $\Sigma_1, B \fCenter \Pi_1, I_2$ and $I_2, \Sigma_2 \fCenter \Pi_2$ are provable.

							Using these prerequisites, we first establish that $I_1 \lor I_2$ fulfills conditions \ref{maehcond1} and \ref{maehcond2} of an interpolant for $\partisym$:
							\medskip

							\begin{prooftree}
								\Axiomm{\Gamma_1 \fCenter \Delta_1, A, I_1}
								\Axiomm{\Sigma_1, B \fCenter \Pi_1, I_2}
								\RightLabelm{\lkrule{\limpl}{l}}
								\BinaryInfm{\Gamma_1, \Sigma_1, A\limpl B \fCenter \Delta_1, \Pi_1, I_1, I_2}
								\RightLabelm{\lkrule{\lor}{r}}
								\UnaryInfm{\Gamma_1, \Sigma_1, A\limpl B \fCenter \Delta_1, \Pi_1, I_1 \lor I_2}
							\end{prooftree}
							\medskip

							\begin{prooftree}
								\Axiomm{I_1, \Gamma_2 \fCenter \Delta_2}
								\Axiomm{I_2, \Sigma_2 \fCenter \Pi_2}
								\RightLabelm{\lkrule{\lor}{l}}
								\BinaryInfm{I_1 \lor I_2, \Gamma_2, \Sigma_2 \fCenter \Delta_2, \Pi_2}
							\end{prooftree}
							\medskip

							{
								%\setlength{\abovedisplayskip}{0pt}
								%\setlength{\belowdisplayskip}{0pt}
								%\setlength{\abovedisplayshortskip}{0pt}
								%\setlength{\belowdisplayshortskip}{0pt}


								To show that also condition \ref{maehcond3} is satisfied, consider that by the induction hypothesis, it holds that:
								\begin{align*}
									\Lang(I_1) &\subseteq \Lang(\Gamma_1, \Delta_1, A) \cap \Lang(\Gamma_2, \Delta_2) \\
									\Lang(I_2) &\subseteq \Lang(\Sigma_1, B, \Pi_1) \cap \Lang(\Sigma_2, \Pi_2)
								\end{align*}\nopagebreak
								Therefore
								\begin{align*}
									\Lang(I_1) \cup \Lang(I_2) &\subseteq
									(\Lang(\Gamma_1, \Delta_1, A) \cap \Lang(\Gamma_2, \Delta_2)) \cup ( \Lang(\Sigma_1, B, \Pi_1) \cap \Lang(\Sigma_2, \Pi_2))  \\
									&\Downarrow \\
									\Lang(I_1) \cup \Lang(I_2) &\subseteq
									(\Lang(\Gamma_1, \Delta_1, A) \cup \Lang(\Sigma_1, B, \Pi_1)) \cap (\Lang(\Gamma_2, \Delta_2) \cup \Lang(\Sigma_2, \Pi_2)) \\
									&\Updownarrow \\
									\Lang(I_1 \lor I_2) &\subseteq \Lang(\Gamma_1, \Sigma_1, A\limpl B, \Delta_1, \Pi_1) \cap \Lang(\Gamma_2, \Sigma_2, \Delta_2, \Pi_2)
								\end{align*}

							}

						\item $\partisym = \parti{\Gamma_1, \Sigma_1}{\Delta_1, \Pi_1}{\Gamma_2, \Sigma_2, A\limpl B}{\Delta_2, \Pi_2}$.
							The argument for this case is similar using $I_1 \land I_2$ as interpolant.
					\end{enumerate}


					For the other binary connectives \lkrule{\land}{l}, \lkrule{\land}{r}, \lkrule{\lor}{l}, \lkrule{\lor}{r} and \lkrule{\limpl}{r}, similar arguments go through, where the interpolant is always either the conjunction or the disjunction of the interpolants of partitions of the preceding sequents.

			\end{itemize}

		\item[\normalfont Quantifier rules.]
			Suppose the property holds for $n$ rule applications and the $(n+1)$th rule application is a quantifier rule.

			\begin{itemize}
				\item The last rule application is an instance of $\lkrule{\forall}{l}$. Then it is of the form:
					\begin{prooftree}
						\Axiomm{\Gamma, A\subst{x/y} \fCenter \Delta}
						\RightLabelm{\lkrule{\forall}{l}}
						\UnaryInfm{\Gamma, \forall x A \fCenter \Delta}
					\end{prooftree}
					Note that since we have excluded function symbols from occurring in the final sequent (and constant symbols are treated as function symbols of arity 0) and
					by completeness there is a proof of the sequent in the language of the sequent, we can assume that no function or constant symbols occur in this proof.
					Hence quantifiers are only instantiated by variables.

					There are two possible partition schemes of $\Gamma, \forall x A \vdash \Delta$:
					\begin{enumerate}
						\item \parti{\Gamma_1, \forall x A}{\Delta_1}{\Gamma_2}{\Delta_2}.
							By the induction hypothesis, there is an interpolant $I$ of the partition $\parti{\Gamma_1, A\subst{x/y}}{\Delta_1}{\Gamma_2}{\Delta_2}$.
							Hence for $I$, 
							$\Gamma_1, A\subst{x/y} \fCenter \Delta_1, I$ and  
							$I, \Gamma_2 \fCenter \Delta_2$ are provable.
							By an application of $\lkrule{\forall}{l}$ to the first sequent we get $\Gamma_1, \forall x A\fCenter \Delta_1, I$, so $I$ satisfies conditions \ref{maehcond1} and \ref{maehcond2} of being an interpolant for $\partisym$.

							In order to show that also $\Lang(I) \subseteq \Lang(\Gamma_1, \forall x A, \Delta_1) \cap \Lang(\Gamma_2, \Delta_2)$, consider that by the induction hypothesis, 
							$\Lang(I) \subseteq \Lang(\Gamma_1, A\subst{x/y}, \Delta_1) \cap \Lang(\Gamma_2, \Delta_2)$.

							As free variables are not considered to be part of the language, $L(\forall x A) = L(A\subst{x/y})$.


						\item \parti{\Gamma_1}{\Delta_1}{\Gamma_2, \forall x A}{\Delta_2}.
							This case can be argued analogously.
					\end{enumerate}

					In the case of \lkrule{\exists}{r}, a similar argument goes through.

				\item The last rule application is an instance of $\lkrule{\forall}{r}$. Then it is of the form:\nopagebreak
					\begin{prooftree}
						\Axiomm{\Gamma\fCenter \Delta, A\subst{x/y} }
						\RightLabelm{\lkrule{\forall}{r}}
						\UnaryInfm{\Gamma\fCenter \Delta, \forall x A }
					\end{prooftree}
					where $y$ does not appear in $\Gamma$, $\Delta$ or $A$.

					There are two possible partition schemes of $\Gamma\vdash \Delta, \forall x A $:
					\begin{enumerate}
						\item $\partisym = \parti{\Gamma_1}{\Delta_1, \forall x A}{\Gamma_2}{\Delta_2}$.
							By the induction hypothesis, there exists an interpolant I of the partition 
							\parti{\Gamma_1}{\Delta_1, A\subst{x/y}}{\Gamma_2}{\Delta_2} of the upper sequent.
							Hence for $I$, 
							$\Gamma_1 \fCenter \Delta_1, A\subst{x/y}, I$ and
							$I, \Gamma_2 \fCenter \Delta_2$ are provable.

						As $y$ does not occur in $\Gamma$ or $\Delta$ and consequently by condition \ref{maehcond3} does not occur in $I$, we may apply the $\lkrule{\forall}{r}$ rule to the former sequent to obtain $\Gamma_1 \fCenter \Delta_1, \forall x A, I$.
							Hence $I$ is an interpolant for $\partisym$ as well.

						\item \parti{\Gamma_1}{\Delta_1}{\Gamma_2}{\Delta_2, \forall x A}.
							This case can be argued analogously.
					\end{enumerate}

					In the case of \lkrule{\exists}{l}, a similar argument goes through.
					\qedhere
			\end{itemize}
			\begin{comment} % i do not explain why this need not be here
			\item[\normalfont Equality rules.]
				Suppose the property holds for $n$ rule applications and the $(n+1)$th rule is an equality rule.

				\begin{itemize}
					\item The last rule application is an instance of $\lkrule{=}{r_1}$. Then it is of the form:
						\begin{prooftree}
							\Axiomm{\Gamma\fCenter \Delta, A\subst{T/t} }
							\Axiomm{\Sigma \fCenter \Pi, s=t}
							\RightLabelm{\lkrule{=}{r_1}}
							\BinaryInfm{\Gamma, \Sigma\fCenter \Delta, \Pi, A\subst{T/s}  }
						\end{prooftree}

						There are two possible partition schemes of $\Gamma, \Sigma \vdash \Delta, \Pi A\subst{T/s} $:
						\begin{enumerate}
							\item $\partisym = \parti{\Gamma_1, \Sigma_1}{\Delta_1, \Pi_1, A\subst{T/s}}{\Gamma_2, \Sigma_2}{\Delta_2, \Pi_2}$.  

								By the induction hypothesis, there is an interpolant $I_1$ for the partition $\parti{\Gamma_1}{\Delta_1, A\subst{T/t}}{\Gamma_2}{\Delta_2}$ of the left upper sequent.
								Hence $\Gamma_1 \fCenter \Delta_1, A\subst{T/t}, I_1$ and $I_1, \Gamma_2 \fCenter \Delta_2$.

								We also get by the induction hypothesis that there is an interpolant $I_2$ for the partition $\parti{\Sigma_1}{\Pi_1, s=t}{\Sigma_2}{\Pi_2}$ of the right upper sequent. Here, we have that 
								$\Sigma_1 \fCenter \Pi_1, s=t, I_2$ and $I_2, \Sigma_2 \fCenter \Pi_2$.

								Now we establish that $I_1 \lor I_2$ is an interpolant for $\partisym$.

								\begin{prooftree}
									\Axiomm{\Gamma_1 \fCenter \Delta_1, A\subst{T/t}, I_1}
									\Axiomm{\Sigma_1 \fCenter \Pi_1, s=t, I_2}
									\RightLabelm{\lkrule{=}{r_2}}
									\BinaryInfm{\Gamma_1, \Sigma_1 \fCenter \Delta_1, \Pi_1, A\subst{T/s}, I_1, I_2}
									\RightLabelm{\lkrule{\lor}{r}}
									\UnaryInfm{\Gamma_1, \Sigma_1 \fCenter \Delta_1, \Pi_1, A\subst{T/s}, I_1 \lor I_2}
								\end{prooftree}

								\begin{prooftree}
									\Axiomm{I_1, \Gamma_2 \fCenter \Delta_2}
									\Axiomm{I_2, \Sigma_2 \fCenter \Pi_2}
									\RightLabelm{\lkrule{\lor}{l}}
									\BinaryInfm{I_1\lor I_2, \Gamma_2, \Sigma_2 \fCenter \Delta_2, \Pi_2}
								\end{prooftree}


								We furthermore get by the induction hypothesis that

								$\Lang(I_1) \subseteq \Lang(\Gamma_1, \Delta_1, A\subst{T/t}) \cap (\Gamma_2, \Delta_2)$

								$\Lang(I_2) \subseteq \Lang(\Sigma_1, \Pi_1, s=t) \cap (\Sigma_2, \Pi_2)$

								$\Lang(I_1 \lor I_2) \subseteq \Lang(\Gamma_1, \Sigma_1, \Delta_1, \Pi_1, A\subst{T/s}) \cap (\Gamma_2 \Sigma_2, \Delta_2, \Pi_2)$
						\end{enumerate}
				\end{itemize}
			\end{comment}
	\end{description}
\end{proof}



