\chapter{Reduction to First-Order Logic without Equality}

A common theme of proofs is to avoid the tedious effort of proving the result from first principles by reducing the problem to one that is easier to solve.
In this instance, we are able to give a reduction for finding interpolants in first-order logic \emph{with} equality to first-order logic \emph{without} equality, where it is simpler to give an appropriate algorithm.
This method is due to Craig (\cite{Craig57linear,Craig57three}).

In order to simplify notation, we shall consider constant symbols to be function symbols of arity $0$ in this section.
The general layout of this approach is the following:
From two sets $\Gamma$ and $\Delta$, where $\Gamma \cup \Delta$ is unsatisfiable, we compute two sets $\Gamma'$ and $\Delta'$ which do not make use of equality but simulate the effects of equality in $\Gamma$ and $\Delta$ via axioms.
In the process of this transformation, also function symbols are replaced by predicate symbols with appropriate axioms to make sure that the behaviour of these function-representing predicates is compatible to the one of actual functions.
Now an interpolant for $\Gamma'$ and $\Delta'$ can be derived using an algorithm that is only capable of handling predicate symbols as all other non-logical symbols have been removed.
Since the additional axioms ensure that the newly added predicate symbols mimic equality and functions respectively, we will see that the occurrences of these predicates in the interpolant can be translated back to occurrences of equality and function symbols in first-order logic with equality in the language of $\Gamma$ and $\Delta$, thereby yielding the originally desired interpolant.


\section{Translation of formulas}

As we shall see in this section, first-order formulas with equality can be transformed into first-order formulas without equality in a way that is satisfiability-preserving, which is sufficient for our purposes.

First, we define axioms in a language with fresh symbols which allows for simulation of equality and functions in first-order logic without equality and function symbols:

\begin{defi}[Translation of languages]
	For a first-order language $\LangSym$ and fresh predicate symbols $E$ and $F_f$ for $f\in \FS(\LangSym)$, 
	\defiemph{$\Trans(\LangSym)$} denotes $(\LangSym \cup \{E\}\cup\{F_f\mid f \in\nolinebreak \FS(\LangSym)\}) \setminus\allowbreak (\{=\nolinebreak \} \cup\nolinebreak \FS(\LangSym))$.
\end{defi}

\begin{defi}[Equality and function axioms]
	For a first-order language $\LangSym$ we define the following axions in $\Trans(\LangSym)$:\nopagebreak
	\begin{align*}
		\FAX(\LangSym) \defeq{}& \smashoperator{\bigcup_{f \in \FS(\LangSym)}}  \forall \bar x \exists y (F_f(\bar x, y) \land (\forall z (F_f(\bar x, z) \limpl E(y, z)))) \\
%		\EAX(A) \defeq{} & \forall x \; x=x~\land  \\
%																		 & \begin{aligned} \bigwedge_{\substack{P \in\PS(A)\cup\\ \{F_i\mid f_i\in \FS(A)\}}} &\forall x_1 \ldots \forall x_{\ar(P)} \forall y_1 \ldots \forall y_{\ar(P)} \\
%																 & (( x_1~=~y_1 \land \ldots \land x_{\ar(P)} = y_{\ar(P)}) \limpl  \\
%																 &  (P(x_1, \ldots, x_{\ar(P)}) \Lra P(y_1, \ldots, y_{\ar(P)}) ) ) \end{aligned} 
		\Refl(P) \defeq{}& \forall x P(x,x)  \\ 
		\Congr(P) \defeq{}& 
%\forall x_1 \ldots \forall x_{\ar(P)} \forall y_1 \ldots \forall y_{\ar(P)} \\
\forall x_1 \forall y_1 \ldots \forall x_{\ar(P)} \forall y_{\ar(P)} 
(( E(x_1,y_1) \land \ldots \land E(x_{\ar(P)},  y_{\ar(P)})) \limpl  \\
 & (P(x_1, \ldots, x_{\ar(P)}) \limpl P(y_1, \ldots, y_{\ar(P)}) )) \\
		\EAX(\LangSym) \defeq{}& \Refl(E) \cup \;\smashoperator{\bigcup_{\substack{P \in\PS(\LangSym)\cup\{E\}\cup\\ \{F_f\mid f\in \FS(\LangSym)\}}}}\; \Congr(P)
	%	\EAX(A) \defeq{} & \forall x E(x,x)\land  \\
	%																	 & \begin{aligned} \bigwedge_{\substack{P \in\PS(A)\cup\{E\}\cup\\ \{F_f\mid f\in \FS(A)\}}} &\forall x_1 \ldots \forall x_{\ar(P)} \forall y_1 \ldots \forall y_{\ar(P)} \\
	%																	 & (( E(x_1,y_1) \land \ldots \land E(x_{\ar(P)},  y_{\ar(P)})) \limpl  \\
	%															 &  (P(x_1, \ldots, x_{\ar(P)}) \Lra P(y_1, \ldots, y_{\ar(P)}) ) )
	%\end{aligned} 
\qedhere
\end{align*}
%For sets of first-order formulas $\Phi$ and $h \in \{\FAX, \EAX\}$, $h(\Phi) \defeq \bigcup_{A\in \Phi} h(A)$ .
%$\FAX$ and $\EAX$ generalise to sets of formulas by elementwise application.
\end{defi}

$\Refl(P)$ will be referred to as reflexivity axiom of $P$, $\Congr(P)$ as congruence axiom of $P$.
As any model of $\EAX(\LangSym)$ requires $\Refl(E)$ and $\Congr(E)$, $E$ is also symmetric and transitive in the model:

\begin{prop}
	\label{prop:equivalence_relation}
	In every model of $\Refl(E)$ and $\Congr(E)$,
	$E$ is an equivalence relation.
\end{prop}
\begin{proof}
	Let $M$ be a model of $\Refl(E)$ and $\Congr(E)$.
	Then $M$ clearly is reflexive.
	Due to $M \entails \Congr(E)$,
	$M \entails \forall x \forall y (E(x, y) \land E(x, x)) \limpl ( E(x, x) \limpl\nolinebreak E(y, x))$.
	As we know that $E$ is reflexive, this simplifies to
	$M \entails \forall x \forall y (E(x, y)\limpl\nolinebreak E(y, x))$, i.e.~$E$ is symmetric in $M$.
	We show the transitivity of $E$ by another instance of $\Congr(E)$: 
	$M \entails \forall x \forall y \forall z ((E(y, x) \land E(y, z)) \limpl ( E(y, y) \limpl\nolinebreak E(x, z)))$,
	As $E$ is reflexive and symmetric, we get that 
	$M \entails \forall x \forall y \forall z ((E(x, y) \land E(y, z)) \limpl\nolinebreak E(x, z))$.
\end{proof}


We continue by defining the translation procedure for formulas:


\begin{defi}[Translation and inverse translation of formulas]
	\label{def:trans}
	Let $A$ be a first-order formula and $E$ and $F_f$ for $f \in \FS(A)$ be fresh predicate symbols.
	Then $\Trans(A)$ is the result of applying the following algorithm to $A$:

	\begin{compactenum}
	\item Replace every occurrence of $s=t$ in $A$ by $E(s, t)$
	\label{def:trans_step1}
	\item As long as there is an occurrence of a function symbol $f$ in $A$:
	\label{def:trans_step2}

		Let $B$ be the atom in which $f$ occurs as outermost symbol of a term.
		Then $B$ is of the form $P(s_1, \ldots, s_{j-1}, f(\bar t),\allowbreak s_{j+1}, \ldots s_m)$.
		Replace $B$ in $A$ by $\exists y (F_f(\bar t, y) \land P(s_1, \ldots, s_{j-1}, y, s_{j+1}, \ldots s_m))$ for a fresh variable $y$.
	\end{compactenum}
	\medskip

	Moreover, let the inverse operation $\TransInv(B)$ for formulas $B$ in the language $\Trans(L(A))$ be defined as the result of applying the following algorithm to $B$:
	\begin{compactenum}
	\item Replace every occurrence of $E(s, t)$ in $B$ by $s=t$.
	\item For every $f \in FS(A)$, replace every occurrence of 
		$\exists y (F_f(\bar t, y) \land P(s_1, \ldots, s_{j-1},\allowbreak y,\allowbreak s_{j+1}, \ldots s_m))$
		in $B$ by $P(s_1, \ldots, s_{j-1},\allowbreak f(\bar t),\allowbreak s_{j+1}, \ldots s_m)$ 
		\label{t_inverse_2}

	\item 
		\label{t_inverse_3}
		For every $f \in FS(A)$, replace every occurrence of 
		$F_f(\bar t, s)$ by $f(\bar t) = s$.
	\end{compactenum}

	For sets of first-order formulas $\Phi$, let $\Trans(\Phi) \defeq \bigcup_{A\in\Phi} \Trans(A)$ and 
$\TransInv(\Phi) \defeq \bigcup_{A\in\Phi} \TransInv(A)$.
\end{defi}

\begin{remark}
	Let $\LangSym$ be a language.
	Step \ref{t_inverse_2} and \ref{t_inverse_3} of $\TransInv$ are both concerned with replacing occurrences of $F_f$ by occurrences of $f$ for $f \in \FS(\LangSym)$, but are relevant in different contexts.

	Step \ref{t_inverse_2} of $\TransInv$ is the precise inverse of step \ref{def:trans_step2} of $\Trans$ in the sense that for any formula $A$, $\TransInv(\Trans(A)) = A$ as we will show in Lemma \ref{lemma:tinv}.
	In this context, step \ref{t_inverse_3} has no effect, as all occurrences of $F_f$ have been introduced by $\Trans(\cdot)$ and are consequently of exactly the form that is handled by step \ref{t_inverse_2}. 
	So the algorithm is in this regard complete even without step \ref{t_inverse_3}.

	On the other hand, if arbitrary formulas in the language $\Trans(\LangSym)$ are given, 
	they in general do not match that pattern and are only translated to $\LangSym$ in step \ref{t_inverse_3}.
	Note that $\TransInv$ without step \ref{t_inverse_2} yields a complete algorithm, as any formula that is handled there can also be processed in step \ref{t_inverse_3}.
	In such a procedure, $\TransInv(\Trans(A))$ and $A$ are in general not syntactically equal for formulas $A$ but only logically equivalent. 
\end{remark}

\begin{lemma}
	\label{lemma:tinv}
	Let $A$ be a first-order formula and $\Phi$ be a set of first-order formulas.
	Then 
	$\TransInv(\Trans(A)) = A$
	and
	$\TransInv(\Trans(\Phi)) = \Phi$
	.
\end{lemma}
\begin{proof}
	Step 1 and 2 in the algorithms $\Trans$ and $\TransInv$ are each concerned with a different set of symbols and therefore do not interfere with each other.
	Moreover, the respective steps in both algorithms are the inverse of each other.
	For step 1, this is immediate and for step 2, consider that all occurrences of $F_f$ for $f \in \FS(A)$ in $\Trans(A)$ have been introduced by $\Trans$ and are consequently of the form
	$\exists y (F_f(\bar t, y) \land P(s_1, \ldots, s_{j-1}, y, s_{j+1}, \ldots s_m))$, which is replaced by 
	$P(s_1, \ldots, s_{j-1},\allowbreak f(\bar t),\allowbreak s_{j+1}, \ldots s_m)$ by $\TransInv$.
	As no occurrences of $F_f$ remain, step 3 of $\TransInv$ leaves the formula unchanged. 
\end{proof}

\begin{defi}[Translation of formulas including axioms]
	For first-order formulas $A$, let $\TransAll(A) \defeq
	\left(\bigwedge_{B\in\FAX(\Lang(A))} B \right) \land
	\left(\bigwedge_{B\in\EAX(\Lang(A))} B \right) \land
	\Trans(A)$ and for sets of first-order formulas $\Phi$, let $\TransAll(\Phi) \defeq \FAX(\Lang(\Phi)) \cup  \EAX(\Lang(\Phi)) \cup \Trans(\Phi)$.
\end{defi}


Note that $\TransAll(A)$ contains neither the equality predicate nor function symbols but additional predicate symbols instead. More formally:

%\begin{defi}[continues=exa:cont]
%	Let $\LangSym$ be a first-order language. 
%	Then $\Trans(\LangSym)$ denotes $(\LangSym\cup \{E\}\cup\{F_f\mid f \in \FS(\LangSym)\}) \setminus(\{=\nolinebreak \} \cup\nolinebreak \FS(\LangSym))$.
%\end{defi}


\begin{lemma}~ 
	\label{lemma:transLang}
	\begin{compactenum}
	\item
		Let $\Phi$ be a set of first-order formulas. Then $\TransAll(\Phi)$ is in the language~$\Trans(\Lang(\Phi))$.
		\label{lemma:transLang1}

	\item 
		If $\Psi$ is in the language $\Trans(\LangSym)$, then $\TransInv(\Psi)$ is in the language~$\LangSym$.
		\label{lemma:transLang2}
	\end{compactenum}
\end{lemma}

\begin{prop}
	\label{prop:transSatEquiv}
	Let $\Phi$ be a set of first-order formulas.
	\begin{compactenum}
		\item If $\Phi$ is satisfiable, then so is $\TransAll(\Phi)$.
			\label{prop:transSatEquiv1}
		\item Let $\LangSym$ be a first-order language and $\Phi$ a set of first-order formulas in the language~$\Trans(\LangSym)$.
			If $\FAX(\LangSym) \cup \EAX(\LangSym) \cup \Phi $ is satisfiable, then so is $\TransInv(\Phi)$.
			\label{prop:transSatEquiv2}
	\end{compactenum}
\end{prop}
\begin{proof}
	Suppose $\Phi$ is satisfiable.
	Let $M$ be a model of $\Phi$.
	We show that $\TransAll(\Phi)$ is satisfiable by extending $M$ to the language $\Lang(\Phi)\cup\{E\}\cup\{F_f\mid f \in \FS(A)\}$ and proving that the extended model satisfies $\TransAll(\Phi)$.

	First, let $M \entails E(s, t)$ if and only if $M \entails s = t$.
	By reflexivity of equality, it follows that $M \entails \Refl(E)$.
	As any predicate, in particular $E$ and $F_f$ for every $f \in \FS(\Phi)$, satisfy the congruence axiom with respect to $=$, by the definition of $E$ in $M$, they satisfy the congruence axiom with respect to $E$.
	Therefore $M$ is a model of $\EAX(\Lang(\Phi))$.

	Second, let $M \entails F_f(\bar x, y)$ if and only if $M \entails f(\bar x) = y$ for all $f \in \FS(\Phi)$. 
	Since $M$ is a model of $\Phi$, it maps every function symbol $f$ to a function, which by definition returns a unique result for every combination of parameters.
	This however is precisely the logical requirement on $F_f$ stated by $\FAX(\Lang(\Phi))$,   
	hence $M$ is a model of $\FAX(\Lang(\Phi))$.

	Lastly, we show that $M \entails \Trans(A)$ for all $A \in \Phi$.
	By the above definition of $E$ in $M$, step $\ref{def:trans_step1}$ of the algorithm in Definition \ref{def:trans} yields a formula that is satisfied by $M$ as it satisfies every formula of $\Phi$.
	For step \ref{def:trans_step2}, suppose $P(s_1, \ldots, s_{j-1}, f(\bar t),\allowbreak s_{j+1}, \ldots s_m)$ does (not) hold under $M$.
	Let $y$ be such that $M \entails f(\bar t)=y$.
	By our definition of $F_f$ under $M$, $M\entails F_f(\bar t, y)$ with this unique $y$.
	Hence $\exists y (F_f(\bar t, y) \land P(s_1, \ldots, s_{j-1}, y, \allowbreak s_{j+1}, \ldots s_m))$ does (not) hold under $M$.


	For 2, suppose $\FAX(\LangSym) \cup \EAX(\LangSym) \cup \Phi$ is satisfiable and let $M$ be a model of it.

	First, note that as $M \entails \EAX(\LangSym)$, by Proposition \ref{prop:equivalence_relation}, $\calI_M(E)$ is an equivalence relation.
	Let $D$ be the domain of $M$.
	We build a model $M'$ whose domain $\domainofmodel{M'}$ is the congruence relation of $\domainofmodel{M}$ modulo $\interpretation{M}(E)$.
	The interpretation $\interpretation{M'}$ of $M'$ is obtained from $\interpretation{M}$ by 
	replacing every occurrence of a domain element $d$ by its respective congruence class with respect to $\interpretation{M}(E)$.
	As $M \entails \EAX(\LangSym)$, $\interpretation{M'}$ satisfies the congruence axioms with respect to every function and predicate symbol, and is therefore well-defined.
	Due to this construction, $M' \entails s = t$ if and only if $M \entails E(s, t)$ for all terms $s$ and $t$.

	%We extend a model $M$ of this set of formulas to a model of $\TransInv(\Phi)$ by extending it from the language $\Trans(\LangSym)$ to include $\FS(\LangSym)$.

	%First, let $M\entails s = t$ if and only if $M\entails E(s, t)$.
	%By Proposition \ref{prop:equivalence_relation} and as $M \entails \EAX(\LangSym)$, $E$ is an equivalence relation in $M$.
	%As these properties directly also apply to $=$ in $M$, equality is adheres to the required axioms in $M$.

	Second, let $M\entails f(\bar t) = s$ if and only if $M\entails F_f(\bar t, s)$ for all $f \in \FS(\LangSym)$.
	As by assumption $M$ is a model of $\FAX(A)$, we know that for every $\bar t$, some $s$ with $M\entails F(\bar t, s)$ exists and is uniquely defined.
	Hence $f$ in $M$ refers to a well-defined function.

	Lastly, to show that $M \entails \TransInv(\Phi)$, 
	consider that the interpretations of the predicates $E$ and $=$ coincide in $M$.
	Furthermore, let $B$ be an occurrence of $\exists y (F_f(\bar t, y) \land P(s_1, \ldots, s_{j-1}, y, s_{j+1}, \ldots s_m))$ for some $f \in \FS(\LangSym)$ in $\Phi$.
	Then by the above definition of $f$ in $M$, we have that $B$ is in $M$ equivalent to $\exists y f(\bar t) = y) \land P(s_1, \ldots, s_{j-1}, y, s_{j+1}, \ldots s_m))$, which due to $f$ being a function is equivalent to 
	$M \entails P(s_1, \ldots, s_{j-1}, f(\bar t), s_{j+1}, \ldots s_m))$.

	Similarly, let $B$ be an occurrence of $F_f(\bar t, s)$ in $\Phi$.
	Then by our above definition of $f$ in $M$, we have that $M \entails f(\bar t) = s$ iff $M \entails B$.
\end{proof}

\begin{corr}
	Let $\Phi$ be a set of first-order formulas.
	Then $\Phi$ is satisfiable if and only if $\TransAll(\Phi)$ is satisfiable.
\end{corr}
\begin{proof}
	The left-to-right direction is directly given in Proposition \ref{prop:transSatEquiv}.
	For the other direction, consider that by Proposition \ref{prop:transSatEquiv}, $\TransInv(\Trans(\Phi))$ is satisfiable, which by Lemma \ref{lemma:tinv} is nothing else than $\Phi$.
\end{proof}



\section{Computation of interpolants}

For the proof of the interpolation theorem by reduction we require an algorithm that operates in first-order logic without equality and function symbols, which we describe in this section.

\begin{remark}
	As the idea of this reduction is to simplify the problem by amongst others not considering function symbols,
	resolution-based methods can not be employed in a direct manner.
	This is because function symbols appear naturally in them as they usually handle existential quantification by means of skolemisation, i.e.~a new function symbol is introduced for every occurrence of an existential quantifier in the scope of a universal quantifier.
	Translating the skolemised formulas to a language without function symbols as described in Definition \ref{def:trans} is of no avail since this translation introduces new existential quantifiers for every function symbol it encounters, necessitating skolemisation yet again.
\end{remark}


\begin{lemma}
	\label{lemma:no_equality_in_proof}
	 %Let $\Gamma$ and $\Delta$ be sets of first-order formulas without the equality symbol and $\Gamma \proves \Delta$ be provable in sequent calculus. 
	 Let $\Gamma$ and $\Delta$ be sets of first-order formulas such that the equality symbol does not occur in them and $\Gamma \proves \Delta$ is provable in sequent calculus. 
	 Then there exists a proof of $\Gamma \proves \Delta$ that does not contain the equality symbol.
\end{lemma}
\begin{proof}
	By the soundness of sequent calculus, we obtain that $\Gamma \entails \Delta$.
	But as sequent calculus without equality rules is complete for first-order logic without equality, there is a proof $\pi$ in this calculus.
	However $\pi$ is obviously also a proof in sequent calculus with equality rules.
\end{proof}

\begin{comment}
	\largered{supported idea: there is a proof in LK with equality, so it is valid. it is true in every model. a calculus without equality is complete for every theorem without equality. hence this calculus without equality has a proof for this formula.}
	Let $\pi$ be a proof of $\Gamma \proves \Delta$.

	Suppose $\pi$ contains an instance of the equality axiom $\proves t = t$ for a term $t$.
	As no equality symbol is contained in the end sequent, there has to be a rule application in $\pi$ which removes either $t=t$.
	Only instances of equality rules or cut are capable of this. 

	Consider the case that an equality rule removed $t=t$.
	As the cases for
	$\lkrule{=}{l_1}$,
	$\lkrule{=}{l_2}$,
	$\lkrule{=}{r_1}$ and
	$\lkrule{=}{r_2}$ are similar, we only consider the case of $\lkrule{=}{l_1}$.
	The proof $\pi'$ leading up to the rule application that removes of the occurrence of $t=t$ is of the form:
	\begin{prooftree}
		\AxiomCm{\varphi}
		\noLine
		\UnaryInfCm{\Gamma, A\occurat{t}{p} \proves \Delta}
		\AxiomCm{\psi}
		\noLine
		\UnaryInfCm{\Sigma \proves \Pi, t=t}
		\RightLabelm{\lkrule{=}{l_1}}
		\BinaryInfm{\Gamma, \Sigma, A\occurat{t}{p} \fCenter \Delta, \Pi }
	\end{prooftree}
	We can replace $\pi'$ in $\pi$ by the following to obtain a proof without an occurrence of the equality symbol:
	\begin{prooftree}
		\AxiomCm{\varphi}
		\noLine
		\UnaryInfCm{\Gamma, A\occurat{t}{p} \proves \Delta}
		\RightLabelm{\lkrule{w}{l}}
		\UnaryInfCm{\Gamma, \Sigma, A\occurat{t}{p} \proves \Delta}
		\RightLabelm{\lkrule{w}{r}}
		\UnaryInfCm{\Gamma, \Sigma, A\occurat{t}{p} \proves \Delta, \Pi}
	\end{prooftree}

	TODO: CUT

	Suppose $\pi$ contains an instance of the axiom $A \proves A$ such that the equality symbol occurs in $A$.
	Then $A$ is of the form $s=t$ for terms $s$ and $t$.
	While the occurrence in the consequent might be eliminated by an equality rule application, due to the subformula property, there is no rule in cut-free sequent calculus such that the occurrence in the antecedent is removed.
	Hence it appears in the final sequent, which contradicts the assumption.

	Suppose $\pi$ contains an instance of \lkrule{w}{l} such that the equality symbolc occurs in the principal formula $A$. 
	This case can be argued similarly as for occurrences of $A$ as antecedent of an axiom $A \proves A$.

	Suppose $\pi$ contains an instance of \lkrule{w}{r} such that the equality symbolc occurs in the principal formula $A$. 
	Then as it does not occur in the end sequent, it is removed by either an instance of an equality rule or the cut rule. 
	Suppose it is removed via an equality rule. We consider the case of  $\lkrule{=}{l_1}$.
	\begin{prooftree}

		\AxiomCm{\Lambda, A\occurat{t}{p} \proves \Theta}

		\AxiomCm{\varphi}
		\noLine
		\UnaryInfCm{\Gamma \proves \Delta}
		\RightLabelm{\lkrule{w}{r}}
		\UnaryInfCm{\Gamma \proves \Delta, s=t}
		\noLine
		\UnaryInfCm{\vdots}
		\noLine
		\UnaryInfCm{\Sigma \proves \Pi, s=t}

		\RightLabelm{\lkrule{=}{l_1}}                        
		\BinaryInfm{\Lambda, \Sigma, A\occurat{s}{p} \fCenter \Theta, \Pi }
	\end{prooftree}



	TODO


	(proof might become longer, but how does that work? don't innessential cuts remain?)

\end{proof}
\end{comment}


We now show that interpolants can be computed by means of a sequent calculus based procedure by Maehara.
It is slightly stronger than the required statement as it allows for interpolants of partitions of sequents:

\begin{defi}[Partition of sequents]
	A {partition} of a sequent $\Gamma \proves \Delta$ is denoted by $\parti{\Gamma_1}{\Delta_1}{\Gamma_2}{\Delta_2}$, where
	$\Gamma_1 \uplus \Gamma_2 = \Gamma$ and 
	$\Delta_1 \uplus \Delta_2 = \Delta$.
\end{defi}

\input{chapters/maehara}

This allows us to state the central theorem of this section:
\begin{thm}
	\label{thm:prop_interpol}
	Let $\Gamma$ and $\Delta$ be sets of closed first-order formulas without equality and function symbols such that $\Gamma \cup \Delta$ is unsatisfiable. Then there is an interpolant for $\Gamma$ and~$\Delta$.
\end{thm}
\begin{proof}
	We show that there is an interpolant for $\Gamma \entails \lnot \Delta$, which
	by Proposition \ref{prop:interpolations_equivalent} proves the theorem.
	By the completeness of cut-free sequent calculus, there is a cut-free proof of $\Gamma \vdash \lnot \Delta$.
	By Lemma \ref{lemma:maehara}, there is an interpolant $I$ for the partition \parti{\Gamma}{}{}{\lnot \Delta}.
	$I$ is the desired interpolant for $\Gamma\entails\lnot\Delta$.

\end{proof}



		\section{Proof by reduction}

		Using the results of the previous sections, we can now give a proof of the interpolation theorem:

		\interpolationRevThm*
		\begin{proof}%[Proof of Theorem \ref{thm:interpolation} (Interpolation)]

			Since $\Gamma \cup \Delta$ is unsatisfiable,
			by Proposition \ref{prop:transSatEquiv}, $\TransAll(\Gamma \cup \Delta)$ is unsatisfiable.
			\begin{align*}
				\TransAll(\Gamma \cup \Delta)\,\semiff~&\{\FAX(\Lang(\Gamma\cup\Delta)), \EAX(\Lang(\Gamma \cup\Delta))\} \cup \Trans(\Gamma \cup \Delta) \\
				\semiff~&\{\FAX(\Lang(\Gamma)\cup\Lang(\Delta)), \EAX(\Lang(\Gamma)\cup\Lang(\Delta))\} \cup \Trans(\Gamma )\cup \Trans(\Delta) \\
				\semiff~&\{\FAX(\Lang(\Gamma)) \land \FAX(\Lang(\Delta)), \EAX(\Lang(\Gamma)) \land \EAX(\Lang(\Delta))\} \cup \Trans(\Gamma) \cup \Trans(\Delta) \\
				\semiff~&\{\FAX(\Lang(\Gamma)),\EAX(\Lang(\Gamma))\} \cup \Trans(\Gamma) \cup \{\FAX(\Lang(\Delta)), \EAX(\Lang(\Delta))\} \cup \Trans(\Delta) \\
				\semiff~&\TransAll(\Gamma) \cup \TransAll(\Delta)
			\end{align*}
			%It follows from Lemma \ref{lemma:trans_transform}, that $\TransAll(\Gamma) \cup \TransAll(\Delta)$is unsatisfiable as well.
			Hence  $\TransAll(\Gamma) \cup \TransAll(\Delta)$ is unsatisfiable as well.
			By Lemma \refsub{lemma:transLang}{lemma:transLang1} $\TransAll(\Gamma)$ and $\TransAll(\Delta)$ contain neither function symbols nor the equality symbol.
			Hence by Theorem \ref{thm:prop_interpol}, there is an interpolant $I$ such that
			\begin{enumerate}
				\item $\TransAll(\Gamma) \entails I$
				\item $\TransAll(\Delta) \entails \lnot I$ 
				\item $\Lang(I) \subseteq \Lang(\TransAll(\Gamma)) \cap \Lang(\TransAll(\Delta))$
					\label{proof:interpolation1_3}
			\end{enumerate}

			We now show that $\TransInv(I)$ is an interpolant for $\Gamma$ and $\Delta$.

			$\TransAll(\Gamma) \entails I$ is equivalent to $\TransAll(\Gamma) \cup \{\lnot I\}$ being unsatisfiable.
			Through the unfolding of $\TransAll(\Gamma)$, we get that 
			$\{\FAX(\Lang(\Gamma)), \EAX(\Lang(\Gamma))\} \cup \Trans(\Gamma) \cup \{\lnot I\}$ is unsatisfiable.
			This set of formulas can now be translated back to the original language with the equality symbol and function symbols. 
			More formally, 
			since $\Lang(\lnot I) \subseteq \Lang(\TransAll(\Gamma))$, we can apply Proposition
			\refsub{prop:transSatEquiv}{prop:transSatEquiv2}
			by considering $\Trans(\Gamma) \cup \{\lnot I\}$ as $\Phi$ to conclude that $\TransInv(\Trans(\Gamma) \cup \{\lnot I\})$ is unsatisfiable. By pulling $\TransInv$ inward and an application of Lemma \ref{lemma:tinv}, we get that $\Gamma \cup \{\TransInv(\lnot I)\} = \Gamma \cup \{\lnot \TransInv(I)\}$ is unsatisfiable. 
			Therefore $\Gamma \entails \Trans^{-1}(I)$.

			For $\Delta$, an analogous argument goes through and so from $\TransAll(\Gamma) \entails \lnot I$ we can deduce that $\Delta \entails \lnot \Trans^{-1}(I)$.

			By item \ref{proof:interpolation1_3}, $I$ is in the language $\Lang(\TransAll(\Gamma)) \cap \Lang(\TransAll(\Delta))$, which by Lemma \refsub{lemma:transLang}{lemma:transLang1} is $\Trans(\Lang(\Gamma)) \cap\nolinebreak \Trans(\Lang(\Delta))$. 
				\newcommand{\somespace}{\;}
				\newcommand{\impconn}[1]{\somespace#1\somespace}
				%\medskip
				%\begin{figure}[H]
			\begin{align*}
				\noalign{$\Trans(\Lang(\Gamma))\cap\nolinebreak \Trans(\Lang(\Delta)) =$} 
				& \Big(\Lang(\Gamma)\impconn\cup \{E\}\impconn\cup\{F_f\mid f \in \FS(\Gamma)\}\Big) \impconn\setminus\Big(\{=\nolinebreak \} \impconn\cup \FS(\Gamma)\Big)\,\impconn\cap \\
				& \Big(\Lang(\Delta)\impconn\cup \{E\}\impconn\cup\{F_f\mid f \in \FS(\Delta)\}\Big) \impconn\setminus\Big(\{=\nolinebreak \} \impconn\cup \FS(\Delta)\Big)\\
				=\,& \Big((\Lang(\Gamma) \cap \Lang(\Delta)) \impconn\cup \{E\} \impconn\cup\{F_f\mid f \in \FS(\Gamma)\cap\FS(\Delta)\}\Big)  \impconn\setminus\Big(\{=\} \impconn\cup \FS(\Gamma) \impconn\cup \FS(\Delta)\Big) \\
				=\,& \Big((\Lang(\Gamma) \cap \Lang(\Delta)) \impconn\cup \{E\} \impconn\cup \{F_f\mid f\in \FS( \Lang(\Gamma) \cap \Lang(\Delta))\} \Big) \impconn\setminus \Big( \{=\} \impconn\cup \FS(\Lang(\Gamma) \cap \Lang(\Delta)) \Big)  \\ 
				=\,& \Trans(\Lang(\Gamma) \cap \Lang(\Delta))
			\end{align*}
				%\end{figure}
				%\vspace{-1em}
			As $I$ is in the language $\Trans(\Lang(\Gamma) \cap \Lang(\Delta))$, by Lemma \refsub{lemma:transLang}{lemma:transLang2}, $\TransInv(I)$ is in the language $\Lang(\Gamma) \cap \Lang(\Delta)$.
		\end{proof}


