
\chapter{The semantic perspective on interpolation}
\label{chap:semantic}

An interesting feature of the interpolation theorem is that it admits a proof, which is distinct from the proof-theoretic ones discussed in the foregoing chapters, as it is purely model-theoretic.
It is based on the joint consistency theorem by Robinson (\cite{robinson1956result}), which we show to be equivalent to the interpolation theorem.
The joint consistency theorem itself was originally presented in \cite{robinson1956result} as a proof of Beth's definability theorem, which is discussed in Section~\ref{sec:beth}.

\section{Joint consistency}
\label{sec:joint_consistency}

The joint consistency theorem is based two notions, which we define now:

\begin{defi}[Consistency]
	A set of formulas $\Gamma$ is consistent if it is not the case that $\Gamma \proves \bot$.
\end{defi}

Note that in classical first-order logic, the notions of consistency and satisfiability coincide.

\begin{defi}[Separability]
	Let $\Gamma$ and $\Delta$ be sets of first-order formulas.
	A formula $A$ in the language $\Lang(\Gamma)\cap \Lang(\Delta)$ is said to \defiemph{separate} $\Gamma$ and $\Delta$ if $\Gamma \entails A$ and $\Delta \entails \lnot A$.
	$\Gamma$ and $\Delta$ are \defiemph{separable} if there exists a formula in the language $\Lang(\Gamma)\cap \Lang(\Delta)$ which separates $\Gamma$ and $\Delta$ and \defiemph{inseparable} otherwise.
\end{defi}

Note that for joint consistency, it is not necessary to require the original sets to be consistent as this is implied by separability:

\begin{lemma}
	\label{lemma:insep_consistent}
	Let $\Gamma$ and $\Delta$ be inseparable sets of first-order formulas. Then $\Gamma$ and $\Delta$ are each consistent.
\end{lemma}
\begin{proof}
	Suppose w.l.o.g.\ that $\Gamma$ is inconsistent. Then $\Gamma \entails \bot$, and as $\Delta \entails \top$, $\bot$ separates $\Gamma$ and $\Delta$.
\end{proof}

The joint consistency theorem shows that if there exists no formula in the language $\Lang(\Gamma)\cap \Lang(\Delta)$ which separates $\Gamma$ and $\Delta$, then there exists no formula in any language which separate $\Gamma$ and $\Delta$ as then, $\Gamma \cup \Delta$ is consistent:

\begin{thm}[Robinson's joint consistency theorem]
	\label{thm:robinson}
	Let $\Gamma$ and $\Delta$ be sets of first-order formulas.
	Then $\Gamma \cup \Delta$ is consistent if and only if $\Gamma$ and $\Delta$ are inseparable.
\end{thm}
The following proof essentially follows \cite{Henkin63} and \cite{chang1990model}.
\begin{proof}
	Suppose that $\Gamma\cup\Delta$ is consistent and let $M$ be a model of it.
	Then clearly for every formula $A$, if $\Gamma \entails A$, then $M \entails A$ as $M \entails \Gamma$.
	But $M \entails \Delta$, hence it can not be the case that $\Delta \entails \lnot A$.

	For the other direction, suppose that $\Gamma$ and $\Delta$ are inseparable.
	We proceed by iteratively constructing two maximal consistent sets of formulas $T$ and $T'$ such that $\Gamma \subset T$ and $\Delta \subset T'$ where $T \cup T'$ is consistent in order to then derive a model of this union, thus establishing the consistency of $\Gamma$ and $\Delta$.

	Let $C = \{c_0, c'_0, c_1, c'_1, \dots\}$ be
	a countably infinite set of fresh constant symbols.
	Let $\mathcal{A}_0, \mathcal{A}_1, \dots$ be an enumeration of all sentences in the language $\Lang(\Gamma) \cup C$
	and $\mathcal{B}_0, \mathcal{B}_1, \dots$ an enumeration of all sentences in the language $\Lang(\Delta) \cup\nolinebreak C$.

	Let $T_0 = \Gamma$ and $T'_0 = \Delta$. 
	We construct
	$T_{i+1}$ 
	from
	$T_{i}$
	by means of the following formation rules:
	%\begin{enumerate}[(1)]
	\begin{enumerate}[~~(1)]
		\item
			\label{theory_construction_1}
			If $T_{i} \cup \{\mathcal{A}_i\}$ and $T'_{i}$ are separable, then $T_{i+1} \defeq{}\, T_i$.
		\item Otherwise:
			\label{theory_construction_2}
			\begin{enumerate}[(2a)]
				\item If $\mathcal{A}_i$ is of the form $\exists x A$, then $T_{i+1} \defeq{}\, T_i \cup \{ \mathcal{A}_i, A\subst{x/c_i} \}$.
			\label{theory_construction_2a}
				\item Otherwise $T_{i+1} \defeq{}\, T_i \cup \{ \mathcal{A}_i \}$.
			\label{theory_construction_2b}
			\end{enumerate}
	\end{enumerate}

	\noindent
	$T'_{i+1}$ is formed in a similar fashion:
	\begin{enumerate}[~~(1$'$)]
		\item
			If $T'_{i} \cup \{\mathcal{B}_i\}$ and $T_{i+1}$ are separable, then $T'_{i+1} \defeq{}\, T'_i$.
		\item
			\begin{samepage}
				Otherwise: 
			\begin{enumerate}[~(2$'$a)]
				\item If $\mathcal{B}_i$ is of the form $\exists x A$, then $T'_{i+1} \defeq{}\, T'_i \cup \{ \mathcal{B}_i, A\subst{x/c'_i} \}$.
				\item Otherwise $T'_{i+1} \defeq{}\, T'_i \cup \{ \mathcal{B}_i \}$.
			\end{enumerate}
			\end{samepage}
	\end{enumerate}

	Now let
	$T = \bigcup_{i\vargeq 0} T_i$
	and
	$T' = \bigcup_{i\vargeq 0} T'_i$.
	We prove properties on $T$ and $T'$ which will be vital for the construction of a model of $T\cup T'$:

	\begin{enumerate}[I.]
		\item
			\label{enum:theories_insep}
			$T_i$ and $T'_i$ are inseparable.

			Suppose to the contrary that $T_i$ and $T'_i$ are separable.
			%Then there exists a formula $A$ in the language $\Lang(\Gamma) \cap \Lang(\Delta)$ such that $\Gamma \entails A$ and $\Delta \entails \lnot A$.
			As $\Gamma$ and $\Delta$ are inseparable by assumption,
			there must be a $j<i$ such that $T_{j}$ and $T'_{j}$ are not separable but $T_{j+1}$ and $T'_{j}$ are,
			or $T_{j+1}$ and $T'_{j}$ are not separable but $T_{j+1}$ and $T'_{j+1}$ are.
			Since these two cases are analogous, we only consider the first. 

			Note that by \ref{theory_construction_1} of the construction procedure, if $T_{j} \cup \{ \mathcal{A}_{j} \}$ and $T'_{j}$ are separable, then $T_{j+1} = T_{j}$. 
			But as we have just witnessed that $T_{j}$ and $T_{j+1}$ are different, $T_{j} \cup \{ \mathcal{A}_{j} \}$ and $T'_{j}$ must be inseparable.
			This however also implies that in the construction procedure, \ref{theory_construction_2b} can not be the case as then, $T_{j+1} = T_{j} \cup \{ \mathcal{A}_{j} \}$ would hold, which contradicts the assumption that $T_{j+1}$ and $T'_j$ are separable.

			Hence \ref{theory_construction_2a} must be the case. Therefore $A_j$ is of the form $\exists x A$ and $T_{j+1} = T_{j} \cup \{ \mathcal{A}_{j}, A\subst{x/c_j} \}$.
			As $T_{j} \cup \{ \mathcal{A}_{j}, A\subst{x/c_j}  \}$ and $T'_j$ are separable, there exists a formula $B$ in the language $\Lang(T_{j} \cup \{ \mathcal{A}_{j}, A\subst{x/c_j}  \}) \cap \Lang(T'_j)$
			such that $ T_{j} \cup \{ \mathcal{A}_{j}, A\subst{x/c_j} \} \entails B$ and $T'_j \entails \lnot B$.
			Since $c_j$ is a fresh variable and therefore is not contained in $\Lang(T'_j)$, $c_j$ does not occur in $B$.
			Hence $B$ is in the language $\Lang(T_{j} \cup \{ \mathcal{A}_{j} \}) \cap \Lang(T'_j)$.
			We conclude by showing that $B$ separates $T_{j} \cup \{ \mathcal{A}_{j} \}$ and $T'_{j}$, which is a contradiction to a previous assumption. 
			In order to do so, it only remains to show that 
			$T_{j} \cup \{ \mathcal{A}_{j} \} \entails B$.


			Let $M$ be a model of $T_{j} \cup \{ \mathcal{A}_{j} \}$ in the language
			$\Lang(T_{j} \cup \{ \mathcal{A}_{j} \})$. Note that $c_j$ is not included in this language as $c_j$ is a fresh variable.
			Since $M\entails \exists x A$, let $d$ be such that $M\entails A\subst{x/d}$.
			Let $M'$ be a model which extends $M$ by interpreting $c_j$ as $d$.
			Then $M' \entails T_{j} \cup \{ \mathcal{A}_{j}, A\subst{x/c_j} \}$. But then $M' \entails B$.
			However as $M$ and $M'$ coincide on the interpretation of the symbols of $\Lang(T_{j} \cup \{ \mathcal{A}_{j} \})$
			and $B$ is in this language, $M\entails B$.


			%we show that $T_{j} \cup \{ \mathcal{A}_{j} \} \entails B$

			%This implies that $T_j \cup \{\mathcal{A}_j\} \entails B$ and
			%that $B$ is in the language $\Lang(T_{j} \cup \{ \mathcal{A}_{j}\}) \cap \Lang(T'_j)$,
			%Then however $B$ separates $T_{j} \cup \{ \mathcal{A}_{j} \}$ and $T'_{j}$, which is a contradiction to a previous assumption. 


			\item 
			\label{enum:theories_consistent}
			$T_i$ and $T'_i$ are consistent.

			Immediate by \ref{enum:theories_insep} and Lemma~\ref{lemma:insep_consistent}.

		\item
			\label{enum:each_max_consistent}
			$T$ and $T'$ are each maximal consistent with respect to $\Lang(\Gamma) \cup C$ and $\Lang(\Delta) \cup C$ respectively.

			We show the result for $T$.
			By~\ref{enum:theories_consistent}, $T$ is consistent.
			Suppose that for some~$i$, $\mathcal{A}_i \not\in T$ and $\lnot\mathcal{A}_i \not\in T$.

			Then in the construction of $T$, case \ref{theory_construction_1} must apply for $\mathcal{A}_i$ as the cases \ref{theory_construction_2a} and \ref{theory_construction_2b} each would add $\mathcal{A}_i$ to $T_{i+1}$ and therefore also to $T$.
			However as \ref{theory_construction_1} applies for $\mathcal{A}_i$, $T_i \cup \{\mathcal{A}_i\}$ and $T'_i$ must be separable.
			As $T_i \subseteq T$, also
			$T \cup \{\mathcal{A}_i\}$ and $T'$ are separable, i.e.\ there exists a formula $B_1$ in the language $\Lang(T\cup\{\mathcal{A}_i\}) \cap \Lang(T') = (\Lang(\Gamma) \cap \Lang(\Delta)) \cup C$ such that
			$T \cup \{\mathcal{A}_i\} \entails B_1$ and $T' \entails \lnot B_1$.
			By the deduction theorem, we also have that \markA{}~$T \entails \mathcal{A}_i \limpl B_1$.

			As we also assume that $\lnot \mathcal{A}_i \not\in T$, by a similar argument, there exists a formula $B_2$ in the language  $(\Lang(\Gamma) \cap \Lang(\Delta)) \cup C$ such that 
			\markB{} $T \entails \lnot \mathcal{A}_i \limpl B_2$ and $T' \entails \lnot B_2$.

			Then however \markA{} and \markB{} entail that in any model, depending on whether $\mathcal{A}_i$ holds in the model, at least one of $B_1$ and $B_2$ holds, i.e.\ $T \entails B_1 \lor B_2$.
			But as neither $B_1$ nor $B_2$ hold in $T'$, we obtain that $T' \entails \lnot (B_1 \lor B_2)$, in effect establishing that $B_1 \lor B_2$ separates $T$ and $T'$, a contradiction to \ref{enum:theories_insep}.


		\item
			\label{enum:intersection_consistent}
			$T \cap T'$ is maximal consistent with respect to $(\Lang(\Gamma) \cap \Lang(\Delta)) \cup C$.

			By~\ref{enum:each_max_consistent}, for every formula $A$ in $(\Lang(\Gamma) \cap \Lang(\Delta)) \cup C$ it holds that either 
			$A \in T$ or $\lnot A \in T$ as well as
			$A \in T'$ or $\lnot A \in T'$. As $T$ and $T'$ are inseparable, either $A \in T$ and $A\in T'$ or otherwise $\lnot A \in T$ and $\lnot A \in T'$.

	\end{enumerate}


	As $T$ is consistent, let $M$ be a model of $T$.
	Due to~\ref{enum:each_max_consistent}, for each term $t$ in $\Lang(\Gamma)\cup C$, $\exists x\, (t = x) \in T$ and hence by~\ref{theory_construction_2a}, there is some $c_i \in C$ such that $t=c_i \in T$.
	Therefore we can find a submodel $N$ of $M$ which as $M$ is in the language $\Lang(\Gamma)\cup C$ such that
	every domain element in $N$ corresponds to a constant symbol in $C$.
	%the domain of $N$ is $\{ \interpretation{M}(c) \mid c \in C\}$, where $\interpretation{M}$ is the interpretation of $M$.
	Models $M'$ of $T'$ allow by a similar reasoning for finding such submodels $N'$ of $M'$.

	As by~\ref{enum:intersection_consistent}, $T$ and $T'$ agree on all formulas of $(\Lang(\Gamma) \cap \Lang(\Delta))\cup C$, 
	we are able to find an isomorphism between the reducts $N$ and $N'$ to their common language.
	Hence we may build a common model $K$ based on $N$ and extending it to $\Lang(\Delta)$ by copying the respective interpretation of $N'$ with regard to the isomorphism.
	%Hence we may build $N'_c$ from $N'$ by exchanging every $c'_i \in C$ by its corresponding $c_j \in C$.
	%Now we see that by extending the $N$ to $\Lang(\Delta)$ by copying the interpretation of $N''$,
	Thus as $N\entails T$ and $N'\entails T'$, $K\entails T \cup T'$, which implies that $\Gamma\cup\Delta$ is consistent.
\end{proof}

\section{Joint consistency and interpolation}

The proof given in the previous section is clearly distinct from the ones in the previous chapters as due to its indirect nature, it does not give rise to a practical algorithm, 
whereas the core idea in each of the other ones is defining an interpolant extraction procedure.

%the other ones each are based on interpolant extraction algorithm.

%Despite the fact that the proof given in the previous section is of a different nature than the ones given in the previous chapters, 
Nevertheless, 
it is easy to see that all of these proofs express equivalent notions.
To that end, let us recall the Interpolation Theorem~\ref{thm:interpolation} in the reverse formulation:

\interpolationRevThm*

\begin{prop}
	Theorem~\ref{thm:robinson} and Theorem~\ref{thm:interpolation} are equivalent.
\end{prop}
\begin{proof}
	It is easy to see that the notion of reverse interpolant and separating formulas coincide.
\end{proof}





